---
title: "Module 01 Lab 01"
author: "HD Sheets"
date: "2024-10-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Module 01 Lab 01

For each problem below,  create the cells needed and enter your solution

# Problem 1

a.) Use seq to create a vector of values x from 10 to 1, decreasing in steps of 1/2

```{r}
x <- seq(10, 1, by = -0.5)
x
```


b.) Find the length of this vector, using an R function
```{r}
length_x <- length(x)
length_x
```


c.) Find the index of the element in x equal to 7.5
```{r}
index_7_5 <- which(x == 7.5)
index_7_5
```


d.) Remove the value 7.5 from the vector x,  so the length decreases by 1
```{r}
x <- x[x != 7.5]
x
```


e.) Sort the vector into both increasing form and decreasing form# Problem 1
```{r}
x_increasing <- sort(x)
x_decreasing <- sort(x, decreasing = TRUE)
x_increasing
x_decreasing
```



# Problem 2

Read the help manual for the function runif()

a.) Generate a vector z of 10 random uniform values between 1 and 5
```{r}
z <- runif(10, min = 1, max = 5)
z
```


b.) Find the minimum and maximum
```{r}
min_z <- min(z)
max_z <- max(z)
min_z
max_z
```


c.) Find the indices of the minimum and maximum

```{r}
index_min_z <- which(z == min_z)
index_max_z <- which(z == max_z)
index_min_z
index_max_z
```


d.) Compute the vector of the square values in z, store this in a vector called
     z2
```{r}
z2 <- z^2
z2
```
     

e.) Find out if any of the squared vectors are less than 4

```{r}
any_less_than_4 <- any(z2 < 4)
any_less_than_4

```


f.) Find out of all the squared vectors are less than 22
```{r}
all_less_than_22 <- all(z2 < 22)
all_less_than_22
```


g.) find out how many  of the squared vectors are greater than 15
      (hint,  try summing the test statement)
```{r}
count_greater_than_15 <- sum(z2 > 15)
count_greater_than_15
```
    

# Problem 3

Read the help manual for rnorm to generate random normal values

a.) create a vector of 9 values, all of which are equal to 3
```{r}
vector_of_threes <- rep(3, 9)
vector_of_threes
```


b.) Put this into a 3 by 3 matrix
```{r}
matrix_b <- matrix(vector_of_threes, nrow = 3, ncol = 3)
matrix_b
```


c.) Label the columns as "a", "b","c", and label the rows 1 to 3
```{r}
colnames(matrix_b) <- c("a", "b", "c")
rownames(matrix_b) <- c("1", "2", "3")
matrix_b
```


d.) create a vector of 12 random normal values
```{r}
set.seed(123)
random_normal_values <- rnorm(12)
random_normal_values
```


e.) put the random normal values into  4 x 3 matrix
```{r}
matrix_e <- matrix(random_normal_values, nrow = 4, ncol = 3)
matrix_e
```


f.) Create a 3 x 3 identity matrix,  you will need to look up how to do this 
     in R,   google search "R identity matrix"
```{r}
identity_matrix <- diag(3)
identity_matrix
```

g.) Multiply the matrix from part b by the identify matrix created in part f
```{r}
result_g <- matrix_b %*% identity_matrix
result_g
```


h.) find the row and column numbers of the largest value in the matrix created in 
     step e
```{r}
max_value_index <- which(matrix_e == max(matrix_e), arr.ind = TRUE)
max_value_index
```
     

i.) What is the average value of the 2nd column of the matrix from step e
```{r}
average_second_column <- mean(matrix_e[, 2])
average_second_column
```


j.) Multiply the matrix from part e by the matrix from part b
```{r}
result_j <- matrix_e %*% matrix_b
result_j
```


# Problem 4

Explain the difference between vectors and lists in R

- Vectors are homogeneous, meaning all elements in a vector must be of the same data type and Lists are heterogeneous, meaning they can contain elements of different types

-how are each created?

Lists are  created using the list() function, vectors are created using the c() function.

-What is different about the two?

Vectors support operations like addition, multiplication, and other mathematical functions and lists do not

-Create a list of the values 1,2,3.  Can you muliple the list by 2? NO
No
```{r}
my_list <- list(1, 2, 3)
my_list


```


# Problem 5

a.) Create a list containing the items   "bob",  "sally", c(1,2,3,4), seq(1,5,by=2), c("jah","nah","hah") in it
```{r}
my_list <- list("bob", "sally", c(1, 2, 3, 4), seq(1, 5, by = 2), c("jah", "nah", "hah"))
my_list
```


b.) How would you reference the word "nah"?
```{r}
word_nah <- my_list[[5]][2]
word_nah
```


c.) How would you refernce the y in "sally"
```{r}
y_in_sally <- substr(my_list[[2]], 5, 5)
y_in_sally
```


d.) Extract the vector 1,2,3,4 and store it in a new variable.  Find the mean
```{r}
vector_1_to_4 <- my_list[[3]]
mean_vector_1_to_4 <- mean(vector_1_to_4)
mean_vector_1_to_4
```


e.) Create a new list from the one in part 1 by removing "bob" and "sally" from the list
```{r}
new_list <- my_list[-c(1, 2)]
new_list
```


# Problem 6

Use ?swiss to look at the built-in data set swiss

a.) What type of data type is this?
```{r}
data_type <- class(swiss)
data_type
```


b.) How many rows and columns does it have (use a function to determine this)
```{r}
num_rows <- nrow(swiss)
num_columns <- ncol(swiss)
num_rows
num_columns
```


c.) Use the summary() and str() functions on this object,  explain what they mean
```{r}
summary(swiss)
str(swiss)
```


d.) Print out the column of Agriculture values, but no other columns
```{r}
agriculture_column <- swiss$Agriculture
agriculture_column
```


e.) Which region has the highest education value?
```{r}
region_highest_education <- rownames(swiss)[which.max(swiss$Education)]
region_highest_education
```


f.) Which region has the lowest infant mortality?
```{r}
region_lowest_infant_mortality <- rownames(swiss)[which.min(swiss$Infant.Mortality)]
region_lowest_infant_mortality
```


# Problem 7

a.) Use file.choose() to get the full name and file path of the example file
sales.csv that is included in the Module 02 content


-Do not use file.choose() in the RMD file,  run it in the console.  Including
file.choose() in an RMD will cause a knit errror

b.)  set infile to the path you found and load the sales.csv file

```{R}
# this is my path name for the download,  change it to the file name you get
# using file.choose()

infile="C:\\Users\\freds\\Downloads\\sales (1).csv"

sales_df=read.csv(infile)



```

c.) Use head(), str() and summary() on this data set.   How many orders are there?
```{r}
head(sales_df)
str(sales_df)
summary(sales_df)
```
```{r}
num_orders <- nrow(sales_df)
num_orders
```


d.) Use the unique() function on the Segment column to find out the number of 
distinct segments there are in the data
```{r}
distinct_segments <- length(unique(sales_df$Segment))
distinct_segments
```


e.) How many different customers are there?
```{r}
num_customers <- length(unique(sales_df$Customer.Name))
num_customers
```


f.) Find the highest profit and the index of the highest profit
```{r}
highest_profit <- max(sales_df$Profit)
index_highest_profit <- which.max(sales_df$Profit)
highest_profit
index_highest_profit
```


g.) What was the name of the customer with the highest profit order?
```{r}
customer_highest_profit <- sales_df$Customer.Name[index_highest_profit]
customer_highest_profit

```

h.) Set the Segment variable to be a factor
```{r}
sales_df$Segment <- as.factor(sales_df$Segment)
```

i.) use the lubridate package to set the Order.Data and Ship.Date varialbe to be
lubridate style date variables
```{r}
library(lubridate)
```
```{r}
sales_df$Order.Date <- mdy(sales_df$Order.Date)
sales_df$Ship.Date <- mdy(sales_df$Ship.Date)
```

j.) compute the shipping delay from each order, using order date and ship data
     what were the longest and shortest shipping delays?
```{r}
sales_df$Shipping.Delay <- as.numeric(difftime(sales_df$Ship.Date, sales_df$Order.Date, units = "days"))
longest_delay <- max(sales_df$Shipping.Delay, na.rm = TRUE)
shortest_delay <- min(sales_df$Shipping.Delay, na.rm = TRUE)
longest_delay
shortest_delay
```

k.) Were orders received on Saturday and Sunday?   Were any shipped on Sunday?

```{r}
received_on_weekend <- any(weekdays(sales_df$Order.Date) %in% c("Saturday", "Sunday"))
shipped_on_sunday <- any(weekdays(sales_df$Ship.Date) == "Sunday")
received_on_weekend
shipped_on_sunday
```


l.)  Convert the Customer.Name to all lower case
```{r}
sales_df$Customer.Name <- tolower(sales_df$Customer.Name)

```


m.) Search the customer names and determine how many contain the name John
```{r}
num_john <- sum(grepl("\\bjohn\\b", sales_df$Customer.Name, ignore.case = TRUE))
num_john
```

n.) How many times is the word "table" in the product names
         -careful,   str_detect() might find "table" in "portable",   also be
         careful with case
```{r}
num_table <- sum(grepl("\\btable\\b", sales_df$Product.Name, ignore.case = TRUE))
num_table
```
         
o.) Use tapply to find the mean profit value, grouped by State
   
      https://www.statology.org/tapply-r/
```{r}
mean_profit_by_state <- tapply(sales_df$Profit, sales_df$State, mean)
mean_profit_by_state
```
    
   
p.) use tapply to find the median profit value, grouped by Segment
```{r}
median_profit_by_segment <- tapply(sales_df$Profit, sales_df$Segment, median)
median_profit_by_segment
```


q.) use table to get the number of orders per segment
```{r}
orders_per_segment <- table(sales_df$Segment)
orders_per_segment
```

